---
title: "GSEA_Cisplatin_PTEC_human/rat compared"
author: "Nienke Kalverboer"
date: "2024-06-06"
output: html_document
---

# Input directory, load packages & load GSEA results 
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# output directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

```

# Load packages & data

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)
library(sysfonts)
library(showtext)

# Add a custom font 
font_add_google("Noto Sans", "noto_sans")
showtext_auto()
```

```{r}
GSEA_results_human = readRDS(file.path(output_dir, "20240618_GSEA_results_human40.rds"))

```

```{r}
GSEA_results_rat = readRDS(file.path(output_dir, "20240618_GSEA_results_rat40.rds"))
```

## Combine results

```{r}
combined_top_GSEA <- bind_rows(GSEA_results_human, GSEA_results_rat)
```

# A combined GSEA graph of human and rat top 10 pathways for each treatment condition

```{r}
p_combined <- ggplot(combined_top_GSEA, aes(x = NES, y = reorder(pathway, NES), color = MEAN_ID_TREATMENT)) +
  geom_point(size = 1, position = position_jitter(width = 0, height = 0)) +
  labs(
    title = "Top 10 Pathways by Absolute NES per Condition",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 4),  
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(limits = c(min(combined_top_GSEA$NES), max(combined_top_GSEA$NES)), expand = c(0.05, 0.05)) +
  guides(color = guide_legend(title = "MEAN_ID_TREATMENT"))


print(p_combined)


```

```{r}
combined_top_GSEA$ID2 <- combined_top_GSEA$MEAN_ID_TREATMENT
```

```{r}
# Separate ID2 to get species seperatly
library(tidyr)
combined_GSEA_ID2 = combined_top_GSEA %>% separate("ID2",
                                                into = c("species", "treatment", "hours", "dose"),
                                                sep = "_") 
```

```{r}
# pathways with >1 treatment condition
multiple_pathways_ID <- combined_GSEA_ID2 %>%
  group_by(pathway) %>%
  filter(n_distinct(species) > 1) %>%
  ungroup()


p_combined_ID <- ggplot(multiple_pathways_ID, aes(x = NES, y = reorder(pathway, NES), color = MEAN_ID_TREATMENT)) +
  geom_point(size = 0.9, position = position_jitter(width = 0, height = 0)) +
  labs(
    title = "Top 10 Pathways by Absolute NES >1 species",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10.5, family = "noto_sans"),  
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(limits = c(min(multiple_pathways_ID$NES), max(multiple_pathways_ID$NES)), expand = c(0.1, 0.1)) +
  guides(color = guide_legend(title = "Treatment condition"))


print(p_combined_ID)
```



# Only pathways that occur with multiple treatment conditions

```{r}
# pathways with >1 treatment condition
multiple_pathways <- combined_top_GSEA %>%
  group_by(pathway) %>%
  filter(n_distinct(MEAN_ID_TREATMENT) > 1) %>%
  ungroup()


p_combined <- ggplot(multiple_pathways, aes(x = NES, y = reorder(pathway, NES), color = MEAN_ID_TREATMENT)) +
  geom_point(size = 0.9, position = position_jitter(width = 0, height = 0)) +
  labs(
    title = "Top 10 Pathways by Absolute NES >1 conditions",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 4, family = "noto_sans"),  
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(limits = c(min(multiple_pathways$NES), max(multiple_pathways$NES)), expand = c(0.1, 0.1)) +
  guides(color = guide_legend(title = "MEAN_ID_TREATMENT"))


print(p_combined)
```


# Only pathways that occur in treatment condition

```{r}
# Select pathways only present in 1 condition
single_pathways <- combined_top_GSEA %>%
  group_by(pathway) %>%
  filter(n_distinct(MEAN_ID_TREATMENT) == 1) %>%
  ungroup()


p_combined_single <- ggplot(single_pathways, aes(x = NES, y = reorder(pathway, NES), color = MEAN_ID_TREATMENT)) +
  geom_point(size = 1, position = position_jitter(width = 0, height = 0)) +
  labs(
    title = "Top 10 Pathways by Absolute NES per Condition",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 4, family = "noto_sans"),  
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(limits = c(min(single_pathways$NES), max(single_pathways$NES)), expand = c(0.05, 0.05)) +
  guides(color = guide_legend(title = "MEAN_ID_TREATMENT"))


print(p_combined_single)
```

# Compared between species (not treatment conditions)

## splitting MEAN_ID_TREATMENT 

```{r}

library(tidyr)
combined_GSEA_hr = combined_top_GSEA %>% separate("MEAN_ID_TREATMENT",
                                                into = c("species", "treatment", "hours", "dose"),
                                                sep = "_") 

```


## Graph 

```{r}

p_combined <- ggplot(combined_GSEA_hr, aes(x = NES, y = reorder(pathway, NES), color = species)) +
  geom_point(size = 1, position = position_jitter(width = 0, height = 0)) +
  labs(
    title = "Top 10 Pathways by Absolute NES per species",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 4, family = "noto_sans"),  
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(limits = c(min(combined_top_GSEA$NES), max(combined_top_GSEA$NES)), expand = c(0.05, 0.05)) +
  guides(color = guide_legend(title = "Species"))


print(p_combined)


```
### graph pathways in multiple species

```{r}


multiple_pathways <- combined_GSEA_hr %>%
  group_by(pathway) %>%
  filter(n_distinct(species) > 1) %>%
  ungroup()


p_combined <- ggplot(multiple_pathways, aes(x = NES, y = reorder(pathway, NES), color = species)) +
  geom_point( position = position_jitter(width = 0, height = 0)) +
  
  labs(
    title = "Top 10 Pathways by Absolute NES per species",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 12, family = "noto_sans", face = "bold"),  
    plot.title = element_text(hjust = 0.5, family = "noto_sans", face = "bold", size = 10)
  ) +
  scale_x_continuous(limits = c(min(multiple_pathways$NES), max(multiple_pathways$NES)), expand = c(0.05, 0.05)) +
  guides(color = guide_legend(title = "Species"))

print(p_combined)
```
