---
title: "BOO2024 - ORA/GSEA_Cisplatin_PTEC_Human"
author: "Imke Bruns (MSc), Steven Kunnen (PhD), Giulia Callegaro (PhD), Hugo van Kessel (Msc), Nienke Kalverboer (BSc)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: default
    highlight: kate
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# Setup {.tabset}
```{r include=FALSE, echo=TRUE, message=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
```


## Load packages
### CRAN
```{r}
# Check if pacman is available and install
if(!require("pacman", quietly = T)){install.packages("pacman")}; library(pacman)

# use packman to install CRAN packages
p_load(tidyverse, data.table, enrichR, fgsea)

```


## Set directories
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# output directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

```


## Metadata output
```{r}
metadata = fread(file.path(input_dir, "20240603_filtered_metadata_processedHuman2.csv"))

```


## DESeq2 output
We name the object `deseq_results`, try to read the data into R yourself.
```{r}
deseq_results = readRDS(file.path(output_dir, "20240603_DESeq2_results_unformatted_human.rds"))

```

## Load functions
Functions can also be loaded from a source script, but for clarity we will include them in the script.
```{r}


runEnrichR = function(genes) {
  enrichR = enrichr(
    genes = genes,
    databases = c(
      "BioCarta_2016",
      "HumanCyc_2016",
      "KEGG_2021_Human",
      "Reactome_2022",
      "WikiPathway_2021_Human",
      "InterPro_Domains_2019",
      "GO_Molecular_Function_2023",
      "GO_Cellular_Component_2023",
      "GO_Biological_Process_2023"
    )
  ) 
  
  enrichR = rbindlist(enrichR, idcol = T)
  names(enrichR)[1:2] = c("database", "source")
  return(enrichR)
}

```


# Rat and Human corresponding genes
```{r}
human_ncbi = fread(input = "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Homo_sapiens.gene_info.gz", 
                   stringsAsFactors = FALSE, 
                   fill = TRUE) %>%
  tibble() %>%
  select("gene_symbol" = Symbol,
         "synonym" = Synonyms,
         "entrez_id" = GeneID,
         "dbxrefs" = dbXrefs,
         "modification_date" = Modification_date) %>%
  distinct() %>%
  mutate(dbxrefs = paste0(dbxrefs, "|")) %>%
  mutate(gene_symbol = gene_symbol,
         entrez_id = as.character(entrez_id),
         ensembl_id = str_extract(string = dbxrefs, "(?<=embl:)(.*?)(?=\\|)"),
         hgnc_id = str_extract(string = dbxrefs, "(?<=HGNC:HGNC:)(.*?)(?=\\|)"),
         mim_id = str_extract(string = dbxrefs, "(?<=MIM:)(.*?)(?=\\|)"),
         alliancegenome_id = str_extract(string = dbxrefs, "(?<=AllianceGenome:HGNC:)(.*?)(?=\\|)")) %>%
  select(gene_symbol, entrez_id, ensembl_id, hgnc_id, mim_id, alliancegenome_id, modification_date) %>%
  mutate(user_date = format(Sys.time(), "%Y%m%d"))
```

```{r include=TRUE, echo=TRUE, message=FALSE}

rat_ncbi = fread(input = "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Rattus_norvegicus.gene_info.gz", 
                 stringsAsFactors = FALSE, 
                 fill = TRUE) %>%
  tibble() %>%
  select("gene_symbol" = Symbol,
         "synonym" = Synonyms,
         "entrez_id" = GeneID,
         "dbxrefs" = dbXrefs,
         "modification_date" = Modification_date) %>%
  distinct() %>%
  mutate(dbxrefs = paste0(dbxrefs, "|")) %>%
  mutate(gene_symbol = gene_symbol,
         entrez_id = as.character(entrez_id),
         rgd_id = str_extract(string = dbxrefs, "(?<=RGD:)(.*?)(?=\\|)"),
         ensembl_id = str_extract(string = dbxrefs, "(?<=embl:)(.*?)(?=\\|)"),
         alliancegenome_id = str_extract(string = dbxrefs, "(?<=AllianceGenome:RGD:)(.*?)(?=\\|)")) %>%
  select(gene_symbol, entrez_id, ensembl_id,rgd_id, alliancegenome_id, modification_date) %>%
  mutate(user_date = format(Sys.time(), "%Y%m%d"))

```



```{r}
human_rgd <- fread(input = "https://download.rgd.mcw.edu/data_release/HUMAN/ORTHOLOGS_HUMAN.txt",
                 stringsAsFactors = FALSE, 
                 fill = TRUE, 
                 skip = 15, 
                 header = "auto", 
                 na.strings = c(NA_character_, ""))
 
human_rgd <- human_rgd[, !duplicated(colnames(human_rgd)), with = F] %>% data.frame() %>% 
  select(HUMAN_GENE_SYMBOL,
         HUMAN_GENE_NCBI_GENE_ID,
         RAT_GENE_SYMBOL,
         RAT_GENE_NCBI_GENE_ID) %>% tibble()
 
# update rat_rgd with the most recent gene_symbols
human_rgd$HUMAN_GENE_SYMBOL <- rat_ncbi$gene_symbol[match(human_rgd$RAT_GENE_NCBI_GENE_ID,rat_ncbi$entrez_id)]
human_rgd$HUMAN_GENE_SYMBOL <-     human_ncbi$gene_symbol[match(human_rgd$HUMAN_GENE_NCBI_GENE_ID,human_ncbi$entrez_id)]
```


```{r}
deseq_results = deseq_results |>
  mutate(HUMAN_GENE_SYMBOL = gene_symbol) |> 
  left_join(human_rgd)
```


```{r}
gmt_hallmarks = gmtPathways("msigdb_v2023.2.Hs_GMTs/h.all.v2023.2.Hs.symbols.gmt")
gmt_wikipathways = gmtPathways("msigdb_v2023.2.Hs_GMTs/c2.cp.wikipathways.v2023.2.Hs.symbols.gmt")
gmt_reactome = gmtPathways("msigdb_v2023.2.Hs_GMTs/c2.cp.reactome.v2023.2.Hs.symbols.gmt")

# gmt_wikipathways = gmtPathways(file.path(
#   input_dir,
#   "GSEA_PATHWAYS",
#   "wikipathways.v2023.2.Hs.symbols.gmt"
# ))

# show the first few pathways, and within those, show the first few genes
gmt_wikipathways %>% head() %>% lapply(head)


```

# Overrepresentation analysis
```{r error=F,warning=F,message=F}

deseq_results_filtered <- deseq_results[!is.na(deseq_results$RAT_GENE_SYMBOL), ]

# We set the threshold for significant genes
padj_threshold <- 0.05

# We perform ORA analysis using EnrichR
ora <- deseq_results_filtered %>%
  filter(padj < padj_threshold) %>%
  nest_by(MEAN_ID_TREATMENT) %>%
  mutate(significant_genes = list(data %>%
                                    arrange(padj) %>%
                                    # slice(1:50) %>%
                                    pull(gene_symbol))) %>%
  mutate(ora = list(
    runEnrichR(genes = significant_genes) %>% filter(Adjusted.P.value < 0.05)
  ))


# print the output
for (i in ora$MEAN_ID_TREATMENT) {
  print(paste0("Top 10 significant ORA for ", i))
  print(
    ora %>%
      filter(MEAN_ID_TREATMENT == i) %>%
      select(MEAN_ID_TREATMENT, ora) %>%
      unnest(cols = c(ora)) %>%
      select(MEAN_ID_TREATMENT, database, source, Adjusted.P.value, Genes) %>%
      arrange(Adjusted.P.value) %>%
      ungroup() %>%
      slice(1:20)
  )
}

```


# Visualization ORA
```{r}
# Load the required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(sysfonts)
library(showtext)

# Add a custom font 
font_add_google("Noto Sans", "noto_sans")
showtext_auto()
```


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Select top 10 lowest P adjusted values for each condition
get_top_adjusted_p <- function(df, n = 10) {
  df %>%
    arrange(Adjusted.P.value) %>%
    slice_head(n = n)
}

# Combine for visualization
combined_top_ora <- ora %>%
  tidyr::unnest(cols = c(ora)) %>%
  group_by(MEAN_ID_TREATMENT) %>%
  do(get_top_adjusted_p(., 10)) %>%
  ungroup()

# Dotplot
p_ora_combined <- ggplot(combined_top_ora, aes(x = MEAN_ID_TREATMENT, y = source, color = Adjusted.P.value)) +
  geom_point() +
  scale_size_continuous(range = c(1, 10)) +
  scale_color_gradientn(colors = c("darkblue", "blue", "cyan", "yellow", "red"), trans = "log") +  
  labs(
    title = "Top 10 Pathways by Lowest Adjusted P-value per Condition (ORA)",
    x = "MEAN_ID_TREATMENT",
    y = "Source",
    size = 1,
    color = "Adjusted P-value"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6, family = "noto_sans"),
    plot.title = element_text(size = 10) 
  )


print(p_ora_combined)
```


# Save output
```{r}
write_rds(combined_top_ora, file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_ORA_results_human.rds"))) 

```
# ORA multiple conditions

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

get_top_adjusted_p <- function(df, n = 10) {
  df %>%
    arrange(Adjusted.P.value) %>%
    slice_head(n = n)
}


combined_top_ora <- ora %>%
  tidyr::unnest(cols = c(ora)) %>%
  group_by(MEAN_ID_TREATMENT) %>%
  do(get_top_adjusted_p(., 10)) %>%
  ungroup()

# Select sources that occur in multiple treatment conditions
multiple_sources <- combined_top_ora %>%
  group_by(source) %>%
  filter(n() > 1) %>%
  ungroup()

p_ora_combined2 <- ggplot(multiple_sources, aes(x = MEAN_ID_TREATMENT, y = source, color = Adjusted.P.value)) +
  geom_point(size = 3) +
  scale_color_gradientn(colors = c("darkblue", "blue", "cyan", "yellow", "red"), trans = "log") +  
  labs(
    title = "Top 10 Pathways by Lowest Adjusted P-value per Condition (ORA)",
    x = "MEAN_ID_TREATMENT",
    y = "Source",
    color = "Adjusted P-value"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 5),
    plot.title = element_text(size = 10)
  )


print(p_ora_combined2)
```



```{r}
# Select top 20 lowest P adjusted values 
top_20_combined <- combined_top_ora %>%
  arrange(Adjusted.P.value) %>%
  slice_head(n = 20)

p_top_20_combined <- ggplot(top_20_combined, aes(x = Adjusted.P.value, y = source, color = MEAN_ID_TREATMENT)) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Top 20 Pathways with Lowest Adjusted P-values Across Conditions (ORA)",
    x = "Adjusted P-value",
    y = "Source",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 6),
    plot.title = element_text(size = 8)
  )


print(p_top_20_combined)
```

# Create a heatmap 
## Create a dataframe with the top 20 for each condition

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)


fix_column_types <- function(df) {
  df %>%
    mutate(
      database = as.character(database),
      source = as.character(source),
      Adjusted.P.value = as.numeric(Adjusted.P.value)
    )
}

ora_heatmap <- ora %>%
  rowwise() %>%
  mutate(ora = list(fix_column_types(ora)))

# Function to get top 20 lowest Adjusted.P.values per condition
get_top_adjusted_p <- function(df, n = 20) {
  df %>%
    arrange(Adjusted.P.value) %>%
    slice_head(n = n)
}

# Combine the top results per condition
combined_top_ora <- ora_heatmap %>%
  tidyr::unnest(cols = c(ora)) %>%
  group_by(MEAN_ID_TREATMENT) %>%
  do(get_top_adjusted_p(., 20)) %>%
  ungroup()

combined_top_ora <- combined_top_ora %>%
  filter(Adjusted.P.value < 0.05)

```

## Transformation of data for heatmap

```{r}
heatmap_data <- combined_top_ora %>%
  select(MEAN_ID_TREATMENT, source, Adjusted.P.value) %>%
  pivot_wider(names_from = MEAN_ID_TREATMENT, values_from = Adjusted.P.value)


heatmap_data[is.na(heatmap_data)] <- 1

# Heatmap matrix
heatmap_matrix <- as.matrix(heatmap_data[,-1])
rownames(heatmap_matrix) <- heatmap_data$source
```

## Make heatmap
```{r}
# Convert matrix to long format for ggplot2
heatmap_long <- as.data.frame(as.table(heatmap_matrix))

# Heatmap
ggplot(heatmap_long, aes(Var2, Var1, fill = Freq)) +
  geom_tile(color = "white") +
   scale_fill_gradientn(colors = c("darkblue", "blue", "green", "yellow", "red"), trans = "log", na.value = "white") +
  
  labs(
    title = "Heatmap of Top 20 Significant Pathways (ORA)",
    x = "Treatment",
    y = "Pathway",
    fill = "Adjusted P-value"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),
    plot.title = element_text(size = 10, hjust = 0.5)
  )
```

### Unsignificant values to white 

```{r}
# Convert matrix to long format for ggplot2
heatmap_long <- as.data.frame(as.table(heatmap_matrix))

# Heatmap
ggplot(heatmap_long, aes(Var2, Var1, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
    colors = c("darkblue", "blue", "cyan", "yellow", "red"),
    trans = "log",
    na.value = "white",
    limits = c(min(heatmap_long$Freq, na.rm = TRUE), 0.05),
    breaks = c(min(heatmap_long$Freq, na.rm = TRUE), 0.05),
    labels = function(x) format(x, scientific = TRUE)
  ) +
  labs(
    title = "Heatmap of Top 20 Significant Pathways (ORA)",
    x = "Treatment",
    y = "Pathway",
    fill = "Adjusted P-value"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6, family = "noto_sans"),
    plot.title = element_text(size = 10, hjust = 0.5)
  )
```

## Heatmap sources > 1 condition

```{r}
# Select--> pathways in multiple conditions
pathway_counts <- combined_top_ora %>%
 count(source, sort = TRUE) %>%
  filter(n > 1) %>%
  select(source)

#Filter--> only selected pathways 
filtered_combined_top_ora <- combined_top_ora %>%
  filter(source %in% pathway_counts$source)

heatmap_data <- filtered_combined_top_ora %>%
  select(MEAN_ID_TREATMENT, source, Adjusted.P.value) %>%
  pivot_wider(names_from = MEAN_ID_TREATMENT, values_from = Adjusted.P.value)

# Replace NA values
heatmap_data[is.na(heatmap_data)] <- 1
#heatmap_data[heatmap_data] <- NA

# Heatmap matrix
heatmap_matrix <- as.matrix(heatmap_data[,-1])
rownames(heatmap_matrix) <- heatmap_data$source

# Convert matrix to long format
heatmap_long <- as.data.frame(as.table(heatmap_matrix))

# Heatmap
ggplot(heatmap_long, aes(Var2, Var1, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
    colors = c("darkblue", "blue", "cyan", "yellow", "red"),
    trans = "log",
    na.value = "white",
    limits = c(min(heatmap_long$Freq, na.rm = TRUE), 0.05),
    breaks = c(min(heatmap_long$Freq, na.rm = TRUE), 0.05),
    labels = function(x) format(x, scientific = TRUE)
  ) +
  labs(
    title = "Heatmap of Significant Pathways (ORA) >1 condition",
    x = "Treatment",
    y = "Pathway",
    fill = "Adjusted P-value"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10, family = "noto_sans"),
    plot.title = element_text(size = 10, hjust = 0.5)
  )
```



# Save top 40 for Human-Rat analysis
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Ensure the ORA data is processed and structured correctly
fix_column_types <- function(df) {
  df %>%
    mutate(
      database = as.character(database),
      source = as.character(source),
      Adjusted.P.value = as.numeric(Adjusted.P.value)
    )
}

ora_40 <- ora %>%
  rowwise() %>%
  mutate(ora = list(fix_column_types(ora)))

# Function to get top 20 lowest Adjusted.P.values per condition
get_top_adjusted_p <- function(df, n = 40) {
  df %>%
    arrange(Adjusted.P.value) %>%
    slice_head(n = n)
}

# Combine the top results per condition
combined_top40_ora <- ora_40 %>%
  tidyr::unnest(cols = c(ora)) %>%
  group_by(MEAN_ID_TREATMENT) %>%
  do(get_top_adjusted_p(., 40)) %>%
  ungroup()

combined_top_ora <- combined_top_ora %>%
  filter(Adjusted.P.value < 0.05)

```

```{r}

write_rds(combined_top40_ora , file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_ORA_results_human40.rds"))) 

```




# Gene set enrichment analysis {.tabset}
## Load gmt files
Gmt files can be downloaded from the msigdb database (https://www.gsea-msigdb.org/gsea/msigdb/). Load the pathway data of interest here using the `gmtPathways` function
```{r}
gmt_hallmarks = gmtPathways("msigdb_v2023.2.Hs_GMTs/h.all.v2023.2.Hs.symbols.gmt")
gmt_wikipathways = gmtPathways("msigdb_v2023.2.Hs_GMTs/c2.cp.wikipathways.v2023.2.Hs.symbols.gmt")
gmt_reactome = gmtPathways("msigdb_v2023.2.Hs_GMTs/c2.cp.reactome.v2023.2.Hs.symbols.gmt")

# show the first few pathways, and within those, show the first few genes
gmt_wikipathways %>% head() %>% lapply(head)
```
## Run fgsea

```{r}

library(tidyverse)
library(ggplot2)
library(fgsea) 

# Combine databases for the analysis
combined_gmt <- c(gmt_wikipathways, gmt_hallmarks, gmt_reactome)

# Separate gene_symbol into gene_symbol and probe_nr
deseq_results_test <- deseq_results_filtered %>%
  separate(gene_symbol, into = c("gene_symbol", "probe_nr"), sep = "_") %>%
  group_by(MEAN_ID_TREATMENT) %>%
  distinct(gene_symbol, .keep_all = TRUE) %>%
  ungroup()

# Perform GSEA analysis
gsea <- deseq_results_test %>%
  nest_by(MEAN_ID_TREATMENT) %>%
  mutate(ranked_genes = list(
    data %>% arrange(stat) %>% select(gene_symbol, stat) %>% deframe()
  )) %>%
  mutate(fgsea_output = list(
    fgsea(pathways = combined_gmt, stats = ranked_genes)
  ))

# Prepare data for plotting
gsea_plot_data <- gsea %>%
  unnest(cols = c(fgsea_output)) %>%
  select(MEAN_ID_TREATMENT, pathway, padj, NES) %>%
  arrange(desc(abs(NES))) %>%
  ungroup() %>%
  group_by(MEAN_ID_TREATMENT) %>%
 
  slice(1:10) %>%  # Select top 10 significant pathways for each treatment
  ungroup()

print(gsea_plot_data)


```

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

# Unnest to select needed columns
gsea_results <- gsea %>%
  unnest(cols = c(fgsea_output)) %>%
  select(MEAN_ID_TREATMENT, pathway, NES) %>%
  ungroup()

# Select top 20 NES pathways
get_top_abs_nes <- function(df, n = 20) {
  df %>%
    mutate(abs_NES = abs(NES)) %>%
    arrange(desc(abs_NES)) %>%
    slice_head(n = n) %>%
    select(-abs_NES)
}

combined_top_results <- gsea_results %>%
  group_by(MEAN_ID_TREATMENT) %>%
  do(get_top_abs_nes(., 20)) %>%
  ungroup()


p_combined <- ggplot(combined_top_results, aes(x = NES, y = reorder(pathway, NES), color = MEAN_ID_TREATMENT)) +
  geom_point(size = 1, position = position_jitter(width = 0, height = 0)) +
  labs(
    title = "Top 10 Pathways by Absolute NES per Condition",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 3),  
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(limits = c(min(combined_top_results$NES), max(combined_top_results$NES)), expand = c(0.05, 0.05)) +
  guides(color = guide_legend(title = "MEAN_ID_TREATMENT"))


print(p_combined)



```

```{r}

library(tidyr)
library(dplyr)
combined_top_results_conc = combined_top_results %>% separate("MEAN_ID_TREATMENT",
                                                into = c("species", "treatment", "hours", "dose"),
                                                sep = "_") 
 combined_top_results_conc =  combined_top_results_conc %>% unite("Treatment", treatment, hours)

 
 combined_top_results_conc <- combined_top_results_conc %>%
  mutate(dose = factor(dose, levels = c("2.5", "10", "50")))
 
p_combined_conc <- ggplot(combined_top_results_conc, aes(x = NES, y = reorder(pathway, NES), color = dose)) +
  geom_point(size = 1, aes(shape = Treatment), position = position_jitter(width = 0, height = 0)) +
  labs(
    title = "Top 20 Pathways by Absolute NES per Condition",
    x = "Normalized Enrichment Score (NES)",
    y = "Pathway",
    color = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 4),  
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(limits = c(min(combined_top_results_conc$NES), max(combined_top_results_conc$NES)), expand = c(0.05, 0.05)) +
   scale_color_manual(values = c("2.5" = "blue", "10" = "darkgreen", "50" = "red")) +
  #scale_color_viridis_c() +
  guides(color = guide_legend(title = "dose"))


print(p_combined_conc)
```


# Create heatmap
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Select needed columns with unnest
gsea_results_heatmap <- gsea %>%
  unnest(cols = c(fgsea_output)) %>%
  select(MEAN_ID_TREATMENT, pathway, NES) %>%
  ungroup()

# Select top 20 NES for each treatment condition
get_top_abs_nes <- function(df, n = 20) {
  df %>%
    mutate(abs_NES = abs(NES)) %>%
    arrange(desc(abs_NES)) %>%
    slice_head(n = n) %>%
    select(-abs_NES)
}

# Combine for visualization
combined_top_results <- gsea_results_heatmap %>%
  group_by(MEAN_ID_TREATMENT) %>%
  do(get_top_abs_nes(., 20)) %>%
  ungroup()

# Transform data to wide format 
heatmap_data_GSEA <- combined_top_results %>%
  select(MEAN_ID_TREATMENT, pathway, NES) %>%
  pivot_wider(names_from = MEAN_ID_TREATMENT, values_from = NES)

# Heatmap matrix
heatmap_matrix_GSEA <- as.matrix(heatmap_data_GSEA[,-1])
rownames(heatmap_matrix_GSEA) <- heatmap_data_GSEA$pathway

# Convert matrix to long format 
heatmap_long_GSEA <- as.data.frame(as.table(heatmap_matrix_GSEA))

# Heatmap
ggplot(heatmap_long_GSEA, aes(Var2, Var1, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
    colors = c("purple4", "blue"  , "cyan", "green", "yellow","orange", "red"),
    na.value = "white"
  ) +
  labs(
    title = "Heatmap of Top 20 Pathways by Absolute NES per Condition",
    x = "Condition",
    y = "Pathway",
    fill = "NES"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6, family = "noto_sans"),
    plot.title = element_text(size = 8, hjust = 0.5)
  )
```

## Multiple conditions

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(sysfonts)
library(showtext)

font_add_google("Noto Sans", "noto_sans")
showtext_auto()

# Filter--> only pathways that occur in multiple conditions
filtered_data <- heatmap_long_GSEA %>%
  na.omit() %>%
  group_by(Var1) %>%
  filter(n_distinct(Var2) > 1) %>%
  ungroup()

# Heatmap
ggplot(filtered_data, aes(Var2, Var1, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
    colors = c("purple4", "blue", "cyan", "green", "yellow", "orange", "red"),
    na.value = "white"
  ) +
  labs(
    title = "Heatmap of Top 20 Pathways by Absolute NES >1 condition",
    x = "Condition",
    y = "Pathway",
    fill = "NES"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6, family = "noto_sans"),
    plot.title = element_text(size = 8, hjust = 0.5)
  )
```



## split MEAN_ID_TREATMENT
```{r}

library(tidyr)
library(dplyr)
heatmap_long_GSEA_split = heatmap_long_GSEA %>% separate("Var2",
                                                into = c("species", "treatment", "hours", "dose"),
                                                sep = "_") 
 heatmap_long_GSEA_split =  heatmap_long_GSEA_split %>% unite("Treatment", treatment, hours)


```

```{r}
# Heatmap
heatmap_long_GSEA_split <- heatmap_long_GSEA_split %>%
  mutate(dose = factor(dose, levels = c("2.5", "10", "50")))
heatmap_long_GSEA_split <- heatmap_long_GSEA_split %>% na.omit()

ggplot(heatmap_long_GSEA_split, aes(dose, Var1, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
   colors = c(  "purple4", "blue"  , "cyan", "green", "yellow","orange", "red"),
    na.value = "white"
  ) +
  labs(
    title = "Heatmap of Top 20 Pathways by Absolute NES per dose",
    x = "Dose",
    y = "Pathway",
    fill = "NES"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6, family = "noto_sans"),
    plot.title = element_text(size = 8, hjust = 0.5)
  )
```

```{r}
# Heatmap
heatmap_long_GSEA_split2 <- heatmap_long_GSEA_split %>%
  mutate(Treatment = factor(Treatment, levels = c("CSP_8hr", "CSP_24hr", "CSP_72hr")))
heatmap_long_GSEA_split2 <- heatmap_long_GSEA_split2 %>% na.omit()

ggplot(heatmap_long_GSEA_split2, aes(Treatment, Var1, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
   colors = c(  "purple4", "blue"  , "cyan", "green", "yellow","orange", "red"),
    na.value = "white"
  ) +
  labs(
    title = "Heatmap of Top 20 Pathways by Absolute NES per timepoint",
    x = "Timepoint",
    y = "Pathway",
    fill = "NES"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6, family = "noto_sans"),
    plot.title = element_text(size = 7, hjust = 0.5)
  )
```

# Save output
```{r}
write_rds(combined_top_results, file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_GSEA_results_human.rds"))) 

```

# Save top 40 for human-rat analysis

```{r}

combined_top40_results <- gsea_results_heatmap %>%
  group_by(MEAN_ID_TREATMENT) %>%
  do(get_top_abs_nes(., 40)) %>%
  ungroup()
```

```{r}
write_rds(combined_top40_results, file.path(output_dir, paste0(gsub( "-", "", Sys.Date()), "_GSEA_results_human40.rds"))) 

```
